Modelling M/EEG data with Bayesian nonparametric multilevel models

### 2025-02-09 brms_meeg.qmd
including some more refs and using t2 splines for timegen data

---
title: "Modelling M/EEG data with Bayesian nonparametric multilevel models"
shorttitle: "Bayesian M/EEG modelling"
author:
  - name: Ladislas Nalborczyk
    corresponding: true
    orcid: 0000-0002-7419-9855
    email: ladislas.nalborczyk@cnrs.fr
    url: https://lnalborczyk.github.io
    affiliations:
      - id: id1
        name: "Aix Marseille Univ, CNRS, LPL"
        address: "5 avenue Pasteur"
        city: "13100 Aix-en-Provence"
        country: France
  - name: Paul Bürkner
    orcid: 0000-0001-5765-8995
    url: https://paulbuerkner.com
    affiliations:
      - name: "TU Dortmund University, Department of Statistics"
author-note:
  disclosures:
    conflict-of-interest: "The authors have no conflicts of interest to disclose."
    # financial-support: "This study was supported by Grant [Grant Number] from [Funding Source]."
abstract: "Time-resolved electrophysoiological measurements such as those offered by magneto- or electro-encephalography (M/EEG) provide a unique window onto neural activity underlying cognitive process and how they unfold over time. Typically, we are interested in testing whether such measures differ across conditions and/or groups. The conventional approach consists in conducting mass-univariate statistics followed by some form of multiplicity correction (e.g., FDR, FWER) or cluster-based inference. However, these cluster-based methods have an important downside: they shift the focus of inference from the timepoint to the cluster level, thus preventing any conclusion to be made about the onset and offset of effects (e.g., differences across conditions). Here, we introduce a novel *model-based approch* for analysing one-dimensional M/EEG timeseries such as ERPs or decoding timecourses and their differences across conditions. This approach relies on Bayesian nonparametric multilevel modelling (multilevel generalised additive models or Gaussian processes), which outputs posterior probabililities of the effect being above chance at every timestep/voxel, while naturally taking into account the temporal dependencies and between-subject variability present in such data."
keywords: [EEG, MEG, generalised additive models, gaussian processes, mixed-effects models, Bayesian statistics, brms]
floatsintext: true
numbered-lines: true
bibliography:
  - meeg_modelling.bib
  - grateful-refs.bib
suppress-title-page: false
link-citations: true
# see https://www.overleaf.com/learn/latex/Font_typefaces
# and systemfonts::system_fonts()
# pdf-engine: xelatex
# monofont: Roboto
mainfont: CMU Serif
fontsize: 10pt
draftfirst: false
draftall: false
a4paper: true
lang: en
toc: true
language:
  citation-last-author-separator: "&"
  email: "email"
  title-block-author-note: "Author note"
  title-block-correspondence-note: "Correspondence concerning this article should be addressed to"
  title-block-role-introduction: "Author roles were classified using the Contributor Role Taxonomy (CRediT; https://credit.niso.org/) as follows:"
format:
  # apaquarto-html: default
  apaquarto-pdf:
    documentmode: doc # man, jou, doc, stu
    keep-tex: true
    include-in-header:
      - text: |
          \usepackage{mathtools}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
# loading packages
library(changepoint)
library(tidyverse)
library(tidybayes)
library(patchwork)
library(MetBrewer)
library(grateful)
library(permuco)
library(scales)
library(pakret)
library(scico)
library(knitr)
library(brms)

@@

Disentangling overlapping processes [@skukies_brain_2024; @skukies_modelling_2021]... Using Bayes factors [@teichmann2022]...

@@

@@

<!--

@@

-->

Following the approach used by @sassenhagen2019 and @rousselet_using_2025, we simulated EEG data stemming from two conditions, one with noise only, and the other with noise + signal. As in previous studies, the noise was generated by superimposing 50 sinusoids at different frequencies, following an EEG-like spectrum [see details and code in @yeung2004]. As in @rousselet_using_2025, the signal was generated from truncated Gaussian with an objective onset at 160 ms, a peak at 250 ms, and an offset at 342 ms. We simulated this signal for 250 timesteps between 0 and 0.5s, akin to & 500 Hz sampling rate. We simulated such data for a group of 20 participants with 50 trials per participant and condition.

```{r eeg-group, echo = FALSE, fig.width = 9, fig.cap = "Some ERPs in two conditions with 50 trials each, for a group of 20 participants."}
# importing R version of Matlab code from Yeung et al. (2004)
source("code/eeg_noise.R")

@@

@@

    # converting results to dataframe
    temp_df <- data.frame(
        x = rep(Xf, 2*nrow(cond1) ),
        y = c(c(t(cond1) ), c(t(cond2) ) ),
        trial = c(rep(1:n_trials, each = length(Xf) ), rep(1:n_trials, each = length(Xf) ) ),
        condition = factor(rep(c("cond1", "cond2"), each = Nf*n_trials) ),
        participant = paste0("participant_", sprintf("%02d", P) )
        ) %>%
        select(participant, condition, trial, x, y)

@@

# plotting the data
raw_df %>%
    summarise(y = mean(y), .by = c(participant, condition, x) ) %>%
    ggplot(aes(x, y, colour = condition) ) +
    geom_hline(yintercept = 0, linetype = 2) +
    geom_vline(xintercept = true_onset, linetype = 2) +
    geom_line(show.legend = FALSE) +
    facet_wrap(~participant) +
    labs(x = "Time (ms)", y = "EEG signal (a.u.)")
```

```{r erp-diff, out.width = "75%", fig.cap = "Group-level average difference between conditions (mean +/- standard error of the mean). The 'true' (population value of the) onset is indicated by the vertical dashed line."}
# averaging across participants
group_df <- raw_df %>%
    pivot_wider(names_from = condition, values_from = y, values_fn = mean) %>%
    mutate(y_diff = cond2 - cond1) %>%
    summarise(
        y_mean = mean(y_diff),
        y_sem = sd(y_diff) / sqrt(n() ),
        .by = x
        ) %>%
    mutate(lower = y_mean - y_sem, upper = y_mean + y_sem)

# plotting the data
group_df %>%
    ggplot(aes(x = x, y = y_mean) ) +
    geom_hline(yintercept = 0, linetype = 2) +
    geom_vline(xintercept = true_onset, linetype = 2) +
    geom_ribbon(
        aes(ymin = lower, ymax = upper, colour = NULL),
        alpha = 0.25, show.legend = FALSE
        ) +
    geom_line(show.legend = FALSE) +
    labs(x = "Time (ms)", y = "EEG difference (a.u.)")
```

@@

Models were fitted using the `brms` package [@brms2017; @brms2018; @nalborczyk2019]...

```{r gam, message = FALSE}
# defining a contrast for condition
contrasts(group_df$condition) <- c(-0.5, 0.5)

# fitting the GAM
gam <- brm(
    # cubic regression splines with k-1 basis functions
    # then we should try s(x, participant, bs = "fs")
    # that is, random factor smooth interactions...
    y ~ condition + s(x, bs = "cr", k = 20, by = condition),
    data = group_df,
    family = gaussian(),
    iter = 5000,
    chains = 4,
    cores = 4,
    file = "models/gam.rds"
    )
```

@@

```{r gp, message = FALSE}
gp_model <- brm(
    # k refers to the number of basis functions for
    # computing Hilbert-space approximate GPs
    y ~ gp(x, k = 20, by = condition),
    # if k = NA (default), exact GPs are computed
    # y ~ gp(x, by = condition),
    data = group_df,
    family = gaussian(),
    control = list(adapt_delta = 0.99),
    iter = 5000,
    chains = 4,
    cores = 4,
    file = "models/gp.rds"
    )
```

@@

```{r gam-preds, echo = FALSE}
# defining a function to plot posterior predictions
plot_post_preds <- function (model, data) {

    post_preds <- conditional_effects(
        x = model,
        effect = "x:condition",
        method = "posterior_epred",
        re_formula = NULL,
        prob = 0.99
        )[[1]]

    post_preds %>%
        ggplot(aes(x = x, y = y, colour = condition, fill = condition) ) +
        geom_hline(yintercept = 0, linetype = 2) +
        geom_point(
            data = data %>% summarise(y = mean(y), .by = c(x, condition) ),
            aes(fill = NULL),
            pch = 21, show.legend = FALSE
            ) +
        geom_line(aes(y = estimate__), show.legend = FALSE) +
        geom_ribbon(
            aes(ymin = lower__, ymax = upper__, colour = NULL),
            alpha = 0.25, show.legend = FALSE
            ) +
        labs(x = "Time (ms)", y = "EEG signal (a.u.)")

@@

```{r gam-gp-preds, fig.width = 8, fig.asp = 0.5, fig.cap = "Posterior predictions of the GAM (left) and GP (right) models."}
# plotting the posterior predictions
plot_post_preds(model = gam, data = group_df) +
    plot_post_preds(model = gp_model, data = group_df)
```

@@

```{r, echo = FALSE, fig.width = 8, fig.asp = 0.5, fig.cap = "Slope for the difference between conditions according to the GAM (left) or the GP (right)."}
# check posterior_smooths()?
# https://www.rdocumentation.org/packages/brms/versions/2.22.0/topics/posterior_smooths.brmsfit
# defining a function to plot posterior slope
plot_post_slope <- function (model, data, eps = 0.1) {

  # defining a sequence of x values to make predictions
  x_seq <- crossing(
      x = seq(min(data$x), max(data$x), length.out = 100),
      condition = c("cond1", "cond2")
      ) %>%
      arrange(condition)

  # retrieving posterior samples
  posterior_samples <- posterior_epred(object = model, newdata = x_seq)

@@

  # computing the probability that y is above 0
  prob_y_above_0 <- data.frame(
      x = seq(min(data$x), max(data$x), length.out = 100)
      ) %>%
      mutate(
          m = colMeans(posterior_samples),
          lower = apply(
              X = posterior_samples,
              MARGIN = 2, FUN = quantile, probs = 0.025
              ),
          upper = apply(
              X = posterior_samples,
              MARGIN = 2, FUN = quantile, probs = 0.975
              )
          )

  # plotting it
  prob_y_above_0 %>%
        ggplot(aes(x = x, y = m) ) +
        geom_hline(yintercept = 0, linetype = 2) +
        geom_line() +
        geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2) +
        # theme_light(base_size = 12) +
        labs(x = "Time (ms)", y = expression(Pr(beta~"|"~data, model) ) )

@@

# using the function with model
plot_post_slope(model = gam, data = group_df, eps = 0.1) +
    plot_post_slope(model = gp_model, data = group_df, eps = 0.1)
```

@@

```{r fig-post-prob-test, echo = FALSE, fig.width = 8, fig.asp = 0.5, fig.cap = "Posterior probability of the EEG signal being above 0+eps according to the GAM (left) and the GP (right)."}
# defining a function to plot posterior prob of slope above 0
plot_post_test <- function (model, data, eps = 0.1) {

    # defining a sequence of x values to make predictions
    x_seq <- crossing(
        x = seq(min(data$x), max(data$x), length.out = 100),
        condition = c("cond1", "cond2")
        ) %>%
        arrange(condition)

    # retrieving posterior samples
    posterior_samples <- posterior_epred(object = model, newdata = x_seq)

@@

    # computing the probability that y is above 0
    prob_y_above_0 <- data.frame(
        x = x_seq$x,
        m = colMeans(posterior_samples > (0 + eps) ),
        lower = apply(
            X = posterior_samples > (0 + eps),
            MARGIN = 2, FUN = quantile, probs = 0.025
            ),
        upper = apply(
            X = posterior_samples > (0 + eps),
            MARGIN = 2, FUN = quantile, probs = 0.975
            )
        )

    # plotting it
    prob_y_above_0 %>%
    ggplot(aes(x = x, y = m) ) +
    geom_hline(yintercept = 0.5, linetype = 2) +
    geom_line() +
    # geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2) +
    # theme_light(base_size = 12) +
    # labs(x = "Time (ms)", y = "Pr(EEG>0)|data)")
    labs(x = "Time (ms)", y = expression(Pr(beta>0~"|"~data, model) ) )

@@

# using the function with the two models
plot_post_test(model = gam, data = group_df, eps = 0.1) +
    plot_post_test(model = gp_model, data = group_df, eps = 0.1)
```

We can also express this as the ratio of posterior probabilities (i.e., $p/(1-p)$) and visualise the timecourse of this ratio superimposed with the conventional BF thresholds (@fig-post-prob-ratio).

```{r fig-post-prob-ratio, echo = FALSE, fig.width = 8, fig.asp = 0.5, fig.cap = "Ratio of posterior probability according to the GAM (left) and the GP (right). Timesteps above threshold (>10) are highlighted in green."}
# defining a sequence of x values to make predictions
n_approx <- 1000
x_seq <- crossing(
    x = seq(min(group_df$x), max(group_df$x), length.out = n_approx),
    condition = c("cond1", "cond2")
    ) %>%
    arrange(condition)

# retrieving posterior samples
posterior_samples <- posterior_epred(object = gam, newdata = x_seq)

@@

# sanity checks
# ts.plot(colMeans(posterior_samples) )
# ts.plot(apply(X = posterior_samples, MARGIN = 2, FUN = mean) )
# ts.plot(apply(X = posterior_samples, MARGIN = 2, FUN = sd) )
# mean(apply(X = posterior_samples, MARGIN = 2, FUN = sd) )

# computing the probability that y is above 0
eps <- 0.05
prob_y_above_0 <- data.frame(
    x = seq(min(group_df$x), max(group_df$x), length.out = n_approx)
    ) %>%
    mutate(post_prob = colMeans(posterior_samples) ) %>%
    mutate(m = colMeans(posterior_samples > eps) ) %>%
    mutate(prob_ratio = m / (1 - m) )

# printing the identified clusters
# the signal was a truncated Gaussian defining an objective onset at 160 ms,
# a maximum at 250 ms, and an offset at 342 ms.
threshold <- 10
exceeding_times_gam <- prob_y_above_0 %>%
    dplyr::filter(prob_ratio > threshold) %>%
    summarise(cluster_onset = min(x), cluster_offset = max(x) ) %>%
    mutate(
        cluster_peak = prob_y_above_0 %>%
            dplyr::filter(post_prob == max(post_prob) ) %>%
            pull(x)
        ) %>%
    mutate(true_onset = 160, true_peak = 250, true_offset = 342) %>%
    mutate(model = "GAM")

# plotting it
gam_ratio_plot <- prob_y_above_0 %>%
    mutate(above_thres = ifelse(prob_ratio > threshold, 1, NA) ) %>%
    ggplot(aes(x = x, y = prob_ratio) ) +
    geom_hline(yintercept = 1, linetype = "dashed") +
    geom_hline(yintercept = 3, linetype = "dashed", color = "darkred") +
    geom_hline(yintercept = 1/3, linetype = "dashed", color = "darkred") +
    geom_hline(yintercept = 10, linetype = "dashed", color = "red") +
    geom_hline(yintercept = 1/10, linetype = "dashed", color = "red") +
    geom_hline(yintercept = 100, linetype = "dashed", color = "orangered") +
    geom_hline(yintercept = 1/100, linetype = "dashed", color = "orangered") +
    geom_line() +
    geom_point(
        data = . %>% dplyr::filter(!is.na(above_thres) ),
        aes(y = threshold),
        colour = "darkgreen",
        shape = 15,
        size = 3,
        na.rm = TRUE
        ) +
    scale_y_log10(labels = label_log(digits = 2), limits = c(NA, 1e4) ) +
    labs(
        x = "Time (ms)",
        y = expression(Pr(beta>0~"|"~data, model) / (1 - Pr(beta>0~"|"~data) ) )
        )

# retrieving posterior samples
posterior_samples <- posterior_epred(object = gp_model, newdata = x_seq)

@@

# computing the probability that y is above 0
eps <- 0.05
prob_y_above_0 <- data.frame(
    x = seq(min(group_df$x), max(group_df$x), length.out = n_approx)
    ) %>%
    mutate(post_prob = colMeans(posterior_samples) ) %>%
    mutate(m = colMeans(posterior_samples > eps) ) %>%
    mutate(prob_ratio = m / (1 - m) )

# printing the identified clusters
# the signal was a truncated Gaussian defining an objective onset at 160 ms,
# a maximum at 250 ms, and an offset at 342 ms.
threshold <- 10
exceeding_times_gp <- prob_y_above_0 %>%
    dplyr::filter(prob_ratio > threshold) %>%
    summarise(cluster_onset = min(x), cluster_offset = max(x) ) %>%
    mutate(
        cluster_peak = prob_y_above_0 %>%
            dplyr::filter(post_prob == max(post_prob) ) %>%
            pull(x)
        ) %>%
    mutate(true_onset = 160, true_peak = 250, true_offset = 342) %>%
    mutate(model = "GP")

>>>>

---
title: "Modelling M/EEG data with Bayesian nonparametric multilevel models"
shorttitle: "Bayesian M/EEG modelling"
author:
  - name: Ladislas Nalborczyk
    corresponding: true
    orcid: 0000-0002-7419-9855
    email: ladislas.nalborczyk@cnrs.fr
    url: https://lnalborczyk.github.io
    affiliations:
      - id: id1
        name: "Aix Marseille Univ, CNRS, LPL"
        address: "5 avenue Pasteur"
        city: "13100 Aix-en-Provence"
        country: France
  - name: Paul Bürkner
    orcid: 0000-0001-5765-8995
    url: https://paulbuerkner.com
    affiliations:
      - name: "TU Dortmund University, Department of Statistics"
author-note:
  disclosures:
    conflict-of-interest: "The authors have no conflicts of interest to disclose."
    # financial-support: "This study was supported by Grant [Grant Number] from [Funding Source]."
abstract: "Time-resolved electrophysiological measurements such as those offered by magneto- or electro-encephalography (M/EEG) provide a unique window onto neural activity underlying cognitive process and how they unfold over time. Typically, we are interested in testing whether such measures differ across conditions and/or groups. The conventional approach consists in conducting mass-univariate statistics followed by some form of multiplicity correction (e.g., FDR, FWER) or cluster-based inference. However, these cluster-based methods have an important downside: they shift the focus of inference from the timepoint to the cluster level, thus preventing any conclusion to be made about the onset and offset of effects (e.g., differences across conditions). Here, we introduce a novel *model-based approch* for analysing one-dimensional M/EEG timeseries such as ERPs or decoding timecourses and their differences across conditions. This approach relies on Bayesian nonparametric multilevel modelling (multilevel generalised additive models or Gaussian processes), which outputs posterior probabililities of the effect being above chance at every timestep/voxel, while naturally taking into account the temporal dependencies and between-subject variability present in such data."
keywords: [EEG, MEG, generalised additive models, gaussian processes, mixed-effects models, Bayesian statistics, brms]
floatsintext: true
numbered-lines: true
bibliography:
  - meeg_modelling.bib
  - grateful-refs.bib
suppress-title-page: false
link-citations: true
# see https://www.overleaf.com/learn/latex/Font_typefaces
# and systemfonts::system_fonts()
# pdf-engine: xelatex
# monofont: Roboto
mainfont: CMU Serif
fontsize: 10pt
draftfirst: false
draftall: false
a4paper: true
lang: en
toc: true
language:
  citation-last-author-separator: "&"
  email: "email"
  title-block-author-note: "Author note"
  title-block-correspondence-note: "Correspondence concerning this article should be addressed to"
  title-block-role-introduction: "Author roles were classified using the Contributor Role Taxonomy (CRediT; https://credit.niso.org/) as follows:"
format:
  # apaquarto-html: default
  apaquarto-pdf:
    documentmode: doc # man, jou, doc, stu
    keep-tex: true
    include-in-header:
      - text: |
          \usepackage{mathtools}
---

```{r setup, include = FALSE}
# loading packages
library(changepoint)
library(tidyverse)
library(tidybayes)
library(patchwork)
library(MetBrewer)
library(grateful)
library(scales)
library(pakret)
library(scico)
library(knitr)
library(brms)

@@

Disentangling overlapping processes [@skukies_brain_2024; @skukies_modelling_2021]... Using Bayes factors [@teichmann2022]... Weighting single trials [@pernet2022]...

See <https://www.let.rug.nl/nerbonne/teach/rema-stats-meth-seminar/presentations/Wieling-2014-GAMs.pdf> and <https://www.let.rug.nl/wieling/Statistics/GAM-EEG/lab/>, <https://www.let.rug.nl/wieling/Statistics/GAM-EEG/GAM-EEG.pdf> [@abugaber2023; @meulman2015]...

Recent example of GLM for EEG [@wüllhorst2025; @fischer2013]... "Similar approaches have been successfully applied to EEG time- (Rousselet et al., 2008) and frequency-domain (Cohen and Cavanagh, 2011) data and allow the simultaneous investigation of multiple independent variables while preserving the high temporal resolution of the EEG."... See also [@rousselet2008; @hauk2006]... Example of two-stage regression analysis [individual-level then group-level, 10.1101/2024.04.10.588861, @dunagan2024]...

@@

See also <https://jtimonen.github.io/lgpr-usage/articles/math.html>...

@@

@@

Following the approach used by @sassenhagen2019 and @rousselet_using_2025, we simulated EEG data stemming from two conditions, one with noise only, and the other with noise + signal. As in previous studies, the noise was generated by superimposing 50 sinusoids at different frequencies, following an EEG-like spectrum [see details and code in @yeung2004]. As in @rousselet_using_2025, the signal was generated from truncated Gaussian with an objective onset at 160 ms, a peak at 250 ms, and an offset at 342 ms. We simulated this signal for 250 timesteps between 0 and 0.5s, akin to a 500 Hz sampling rate. We simulated such data for a group of 20 participants with 50 trials per participant and condition (@fig-eeg).

```{r fig-eeg, echo = FALSE, fig.width = 9, fig.cap = "Some ERPs in two conditions with 50 trials each, for a group of 20 participants."}
# importing R version of Matlab code from Yeung et al. (2004)
source("code/eeg_noise.R")

@@

# removing raw_df
if (exists("raw_df") ) rm(raw_df)

@@

    # converting results to dataframe
    temp_df <- data.frame(
        x = rep(Xf, 2*nrow(cond1) ),
        y = c(c(t(cond1) ), c(t(cond2) ) ),
        trial = c(rep(1:n_trials, each = length(Xf) ), rep(1:n_trials, each = length(Xf) ) ),
        condition = factor(rep(c("cond1", "cond2"), each = Nf * n_trials) ),
        participant = paste0("participant_", sprintf("%02d", P) )
        ) %>%
        select(participant, condition, trial, x, y)

@@

# converting time from ms to seconds
raw_df <- raw_df %>%
    # converting time from ms to seconds
    mutate(x = x / 1000) %>%
    # renaming columns
    rename(time = x, eeg = y)

# defining the true onset and offset in seconds
true_onset <- 0.160
true_offset <- 0.342

# plotting the data
raw_df %>%
    summarise(eeg = mean(eeg), .by = c(participant, condition, time) ) %>%
    ggplot(aes(x = time, y = eeg, colour = condition) ) +
    geom_hline(yintercept = 0, linetype = 2) +
    geom_vline(xintercept = true_onset, linetype = 2) +
    geom_line(show.legend = FALSE) +
    facet_wrap(~participant) +
    labs(x = "Time (s)", y = "EEG signal (a.u.)")
```

We computed the average of the ERP difference (@fig-erp)...

```{r fig-erp, echo = FALSE, out.width = "75%", fig.cap = "Group-level average difference between conditions (mean +/- standard error of the mean). The 'true' (population value of the) onset is indicated by the vertical dashed line."}
# averaging across participants
raw_df %>%
    pivot_wider(names_from = condition, values_from = eeg, values_fn = mean) %>%
    mutate(eeg_diff = cond2 - cond1) %>%
    summarise(
        eeg_mean = mean(eeg_diff),
        eeg_sem = sd(eeg_diff) / sqrt(n() ),
        .by = time
        ) %>%
    mutate(lower = eeg_mean - eeg_sem, upper = eeg_mean + eeg_sem) %>%
    # plotting the data
    ggplot(aes(x = time, y = eeg_mean) ) +
    geom_hline(yintercept = 0, linetype = 2) +
    geom_vline(xintercept = true_onset, linetype = 2) +
    geom_ribbon(
        aes(ymin = lower, ymax = upper, colour = NULL),
        alpha = 0.25, show.legend = FALSE
        ) +
    geom_line(show.legend = FALSE) +
    labs(x = "Time (s)", y = "EEG difference (a.u.)")
```

@@

We then fitted a GAM and GP regression model using the `brms` package [@brms2017; @brms2018; @nalborczyk2019]...

```{r gam, message = FALSE}
# averaging across participants
ppt_df <- raw_df %>%
    group_by(participant, condition, time) %>%
    summarise(eeg = mean(eeg) ) %>%
    # mutate(eeg = as.numeric(scale(x = eeg, scale = FALSE) ) ) %>%
    ungroup()

# defining a contrast for condition
contrasts(ppt_df$condition) <- c(-0.5, 0.5)

# fitting the GAM
gam <- brm(
    # cubic regression splines with k-1 basis functions
    eeg ~ condition + s(time, bs = "cr", k = 10, by = condition),
    data = ppt_df,
    family = gaussian(),
    iter = 5000,
    chains = 4,
    cores = 4,
    file = "models/gam.rds"
    )
```

@@

```{r gp, message = FALSE}
# computing eeg_diff and averaging across participants
# group_df <- raw_df %>%
#     group_by(participant, time) %>%
#     pivot_wider(names_from = condition, values_from = eeg, values_fn = mean) %>%
#     mutate(eeg_diff = cond2 - cond1) %>%
#     summarise(eeg_diff = mean(eeg_diff) ) %>%
#     ungroup()

# see https://betanalpha.github.io/assets/case_studies/gp_part3/part3.html
# and https://betanalpha.github.io/assets/case_studies/gaussian_processes.html
gp_priors <- c(
    set_prior("normal(0, 1)", class = "Intercept"),
    set_prior("normal(0, 1)", class = "b"),
    set_prior("exponential(1)", class = "sigma"),
    set_prior("exponential(1)", class = "sdgp")
    )

gp_model <- brm(
    # k refers to the number of basis functions for
    # computing Hilbert-space approximate GPs
    # if k = NA (default), exact GPs are computed
    # eeg ~ gp(time, by = condition),
    eeg ~ condition + gp(time, by = condition, k = 20, cov = "exp_quad"),
    data = ppt_df,
    family = gaussian(),
    # prior = gp_priors,
    control = list(adapt_delta = 0.99, max_treedepth = 15),
    backend = "cmdstanr",
    iter = 2000,
    chains = 4,
    cores = 4,
    file = "models/gp.rds"
    )
```

@@

```{r gam-preds, echo = FALSE}
# defining a function to plot posterior predictions
plot_post_preds <- function (model, data = NULL) {

    if (is.null(data) ) data <- model$data

    post_preds <- conditional_effects(
        x = model,
        effect = "time:condition",
        method = "posterior_epred",
        re_formula = NULL,
        prob = 0.99
        )[[1]]

    post_preds %>%
        ggplot(aes(x = time, y = eeg, colour = condition, fill = condition) ) +
        geom_vline(xintercept = true_onset, linetype = 2) +
        geom_hline(yintercept = 0, linetype = 2) +
        geom_point(
            data = data %>% summarise(eeg = mean(eeg), .by = c(time, condition) ),
            aes(fill = NULL),
            pch = 21, show.legend = FALSE
            ) +
        geom_line(aes(y = estimate__), show.legend = FALSE) +
        geom_ribbon(
            aes(ymin = lower__, ymax = upper__, colour = NULL),
            alpha = 0.25, show.legend = FALSE
            ) +
        labs(x = "Time (s)", y = "EEG signal (a.u.)")

@@

```{r gam-gp-preds, fig.width = 8, fig.asp = 0.5, fig.cap = "Posterior predictions of the GAM (left) and GP (right) models."}
# plotting the posterior predictions
plot_post_preds(model = gam) + plot_post_preds(model = gp_model)
```

@@

```{r, echo = FALSE, fig.width = 8, fig.asp = 0.5, fig.cap = "Slope for the difference between conditions according to the GAM (left) or the GP (right)."}
# check posterior_smooths()?
# https://www.rdocumentation.org/packages/brms/versions/2.22.0/topics/posterior_smooths.brmsfit
# defining a function to plot posterior slope
plot_post_slope <- function (model, data = NULL, eps = 0.1) {

    # if no data is specified, use model$data
    if (is.null(data) ) data <- model$data

    # defining a sequence of time values to make predictions
    time_seq <- crossing(
        time = seq(min(data$time), max(data$time), length.out = 100),
        condition = c("cond1", "cond2")
        ) %>%
        arrange(condition)

    # retrieving posterior samples
    posterior_samples <- posterior_epred(object = model, newdata = time_seq)

@@

    # computing the probability that y is above 0
    prob_y_above_0 <- data.frame(
        time = seq(min(data$time), max(data$time), length.out = 100)
        ) %>%
        mutate(
          m = colMeans(posterior_samples),
          lower = apply(
              X = posterior_samples,
              MARGIN = 2, FUN = quantile, probs = 0.025
              ),
          upper = apply(
              X = posterior_samples,
              MARGIN = 2, FUN = quantile, probs = 0.975
              )
          )

    # plotting it
    prob_y_above_0 %>%
        ggplot(aes(x = time, y = m) ) +
        geom_hline(yintercept = 0, linetype = 2) +
        geom_vline(xintercept = true_onset, linetype = 2) +
        geom_line() +
        geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2) +
        labs(x = "Time (s)", y = expression(Pr(beta~"|"~data, model) ) )

@@

# using the function with model
plot_post_slope(model = gam, eps = 0.1) +
    plot_post_slope(model = gp_model, eps = 0.1)
```

@@

```{r fig-post-prob-test, echo = FALSE, fig.width = 8, fig.asp = 0.5, fig.cap = "Posterior probability of the EEG signal being above 0+eps according to the GAM (left) and the GP (right)."}
# defining a function to plot posterior prob of slope above 0
plot_post_test <- function (model, data = NULL, eps = 0.1) {

    # if no data is specified, use model$data
    if (is.null(data) ) data <- model$data

    # defining a sequence of time values to make predictions
    time_seq <- crossing(
        time = seq(min(data$time), max(data$time), length.out = 100),
        condition = c("cond1", "cond2")
        ) %>%
        arrange(condition)

    # retrieving posterior samples
    posterior_samples <- posterior_epred(object = model, newdata = time_seq)

@@

    # computing the probability that y is above 0
    prob_y_above_0 <- data.frame(
        time = time_seq$time,
        m = colMeans(posterior_samples > (0 + eps) ),
        lower = apply(
            X = posterior_samples > (0 + eps),
            MARGIN = 2, FUN = quantile, probs = 0.025
            ),
        upper = apply(
            X = posterior_samples > (0 + eps),
            MARGIN = 2, FUN = quantile, probs = 0.975
            )
        )

    # plotting it
    prob_y_above_0 %>%
    ggplot(aes(x = time, y = m) ) +
    geom_hline(yintercept = 0.5, linetype = 2) +
    geom_vline(xintercept = true_onset, linetype = 2) +
    geom_vline(xintercept = true_offset, linetype = 2) +
    geom_line() +
    labs(x = "Time (s)", y = expression(Pr(beta>0~"|"~data, model) ) )

@@

# using the function with the two models
plot_post_test(model = gam, eps = 0.1) +
    plot_post_test(model = gp_model, eps = 0.1)
```

We can also express this as the ratio of posterior probabilities (i.e., $p/(1-p)$) and visualise the timecourse of this ratio superimposed with the conventional BF thresholds (@fig-post-prob-ratio). Also check this paper (<https://link.springer.com/article/10.3758/s13428-021-01613-6?fromPaywallRec=false>)?

```{r fig-post-prob-ratio, echo = FALSE, fig.width = 8, fig.asp = 0.5, fig.cap = "Ratio of posterior probability according to the GAM (left) and the GP (right). Timesteps above threshold (>10) are highlighted in green."}
# defining a sequence of x values to make predictions
n_approx <- 1000
time_seq <- crossing(
    time = seq(min(ppt_df$time), max(ppt_df$time), length.out = n_approx),
    condition = c("cond1", "cond2")
    ) %>%
    arrange(condition)

# retrieving posterior samples
posterior_samples <- posterior_epred(object = gam, newdata = time_seq)

@@

# computing the probability that y is above 0
eps <- 0.05
prob_y_above_0 <- data.frame(
    time = seq(min(ppt_df$time), max(ppt_df$time), length.out = n_approx)
    ) %>%
    mutate(post_prob = colMeans(posterior_samples) ) %>%
    mutate(m = colMeans(posterior_samples > eps) ) %>%
    mutate(prob_ratio = m / (1 - m) )

# printing the identified clusters
# the signal was a truncated Gaussian defining an objective onset at 160 ms,
# a maximum at 250 ms, and an offset at 342 ms.
threshold <- 10
exceeding_times_gam <- prob_y_above_0 %>%
    dplyr::filter(prob_ratio > threshold) %>%
    summarise(cluster_onset = min(time), cluster_offset = max(time) ) %>%
    mutate(
        cluster_peak = prob_y_above_0 %>%
            dplyr::filter(post_prob == max(post_prob) ) %>%
            pull(time)
        ) %>%
    mutate(true_onset = 0.160, true_peak = 0.250, true_offset = 0.342) %>%
    mutate(model = "GAM")

# plotting it
gam_ratio_plot <- prob_y_above_0 %>%
    mutate(above_thres = ifelse(prob_ratio > threshold, 1, NA) ) %>%
    ggplot(aes(x = time, y = prob_ratio) ) +
    geom_hline(yintercept = 1, linetype = "dashed") +
    geom_hline(yintercept = 3, linetype = "dashed", color = "darkred") +
    geom_hline(yintercept = 1/3, linetype = "dashed", color = "darkred") +
    geom_hline(yintercept = 10, linetype = "dashed", color = "red") +
    geom_hline(yintercept = 1/10, linetype = "dashed", color = "red") +
    geom_hline(yintercept = 100, linetype = "dashed", color = "orangered") +
    geom_hline(yintercept = 1/100, linetype = "dashed", color = "orangered") +
    geom_vline(xintercept = true_onset, linetype = 2) +
    geom_vline(xintercept = true_offset, linetype = 2) +
    geom_line() +
    geom_point(
        data = . %>% dplyr::filter(!is.na(above_thres) ),
        aes(y = threshold),
        colour = "darkgreen",
        shape = 15,
        size = 3,
        na.rm = TRUE
        ) +
    scale_y_log10(labels = label_log(digits = 2), limits = c(NA, 1e4) ) +
    labs(
        x = "Time (s)",
        y = expression(Pr(beta>0~"|"~data) / (1 - Pr(beta>0~"|"~data) ) )
        )

# retrieving posterior samples
posterior_samples <- posterior_epred(object = gp_model, newdata = time_seq)

@@

# computing the probability that y is above 0
prob_y_above_0 <- data.frame(
    time = seq(min(ppt_df$time), max(ppt_df$time), length.out = n_approx)
    ) %>%
    mutate(post_prob = colMeans(posterior_samples) ) %>%
    mutate(m = colMeans(posterior_samples > eps) ) %>%
    mutate(prob_ratio = m / (1 - m) )

# printing the identified clusters
# the signal was a truncated Gaussian defining an objective onset at 160 ms,
# a maximum at 250 ms, and an offset at 342 ms
exceeding_times_gp <- prob_y_above_0 %>%
    dplyr::filter(prob_ratio > threshold) %>%
    summarise(cluster_onset = min(time), cluster_offset = max(time) ) %>%
    mutate(
        cluster_peak = prob_y_above_0 %>%
            dplyr::filter(post_prob == max(post_prob) ) %>%
            pull(time)
        ) %>%
    mutate(true_onset = 0.160, true_peak = 0.250, true_offset = 0.342) %>%
    mutate(model = "GP")
